Coalescent inference of mutation spectrum histories from sample frequency spectra

### 2019-08-23 manuscript/main.tex
correcting horizon-aware Tinkhonov matrix (tex and python)

Following \cite{Rosen2018-bb} (appendix proof of Prop.\ (1)) let $R_\eta(t) \equiv \int_0^t\frac{dx}{\eta(x)}$, and substitute $\tau \equiv R_\eta(t)$ in \ref{eqn:d} to give
\begin{equation}
\label{eqn:d2}
d_j = \int_0^\infty \exp\left[-\binom{j}{2}\tau\right] \tilde\eta(\tau)\tilde\mu(\tau)dt,
\end{equation}
where $\tilde\eta(\tau) \equiv \eta(R^{-1}(\tau))$ and $\tilde\mu(\tau) \equiv \mu(R^{-1}(\tau))$.
We consider piecewise constant $\eta$ and $\mu$ on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:model:xi}).
We take the epochs as fixed parameters, and in practice make them dense so as to approximate infinite dimensional histories.
Let $(y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $(z_1,\dots,z_m)$ denote the constant mutation rate $\mu(t)$ during each epoch.
Let $u_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=1,\dots,m$.%, and $u_0\equiv 1$.
With this we can follow the proof of Prop.\ (1) in \cite{Rosen2018-bb} mutatis mutandis to arrive at
\begin{equation}
\label{eqn:d3}
\boldsymbol d = M(\boldsymbol y) \boldsymbol z
\end{equation}
where
\begin{equation}
\label{eqn:M}
M(\boldsymbol y) \equiv
\begin{bmatrix}
1 &             &        &                       \\
  & \frac{1}{3} &        &                       \\
  &             & \ddots &                       \\
  &             &        & \frac{1}{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
1       & u_1                & \hdots & \prod_{i=1}^{m-1}u_i               \\
1       & u_1^3              & \hdots & \prod_{i=1}^{m-1}u_i^3             \\
\vdots  & \vdots             & \ddots & \vdots                                \\
1       & u_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{m-1}u_i^{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
1  &      &        &             &       \\
-1 & 1  &        &             &       \\
     & -1 & 1    &             &       \\
     &      & \ddots & \ddots      &       \\
     &      &        & -1 & 1
\end{bmatrix}
\begin{bmatrix}
y_1 &     &      &             &       \\
    & y_2 &      &             &       \\
     &     & y_3 &             &       \\
     &      &    & \ddots      &       \\
     &      &        &  & y_m
\end{bmatrix}.
\end{equation}
Note that the $(n-1)\times m$ matrix $M(\boldsymbol y)$ is a nonlinear function of the demographic history $\boldsymbol y$ because the $u_l$ are nonlinear functions of $\boldsymbol y$.

@@

@@

The inverse problem \eqref{eqn:xivec2} is ill-posed in general, so many very different histories can be equally consistent with the data \citep{oscillation paper? Yun's other papers?}.
To penalize oscillatory solutions we introduce regularizers on the history derivatives.
Intuitively, we know the SFS can't contain any information about the history beyond the TMRCA $T_2$, since mutations that occurred before then will not be segregating in the sample.
Thus, we will find it useful to penalize complexity in the history more heavily at times that are more probably ancestral to the TMRCA.
To do this we should like to use the CDF of the TMRCA to define a ramp of weights.
At times for which the CDF is near unity, the TMRCA has probably occurred, so we should penalize heavily because the data are not informative about the history at this time.
At times for which the CDF is small, the TMRCA has probably not yet occurred, and we should penalize less heavily.
From \eqref{eqn:pi} and \eqref{eqn:r} the CDF is
\begin{align}
F_2(t) &= 1 - \sum_{j=2}^n A_{2,j}r_2(t)\\
&= 1 - \sum_{j=2}^n A_{2,j}r_2(t)
\end{align}
We define $L_2$ derivative regularizers for arbitrary histories $\eta(t)$ and $\mu(t)$ to have unit weight when the CDF is zero, and approach weight $h$ as the CDF approaches unity:
\begin{align}
\rho_\eta[\eta(t); h] &\equiv \int_0^\infty\left(\frac{d\eta(t)}{dt}\right)^2\left[1+(h-1)F_2(t)\right]dt,\\
\rho_\mu[\eta(t), \mu(t); h] &\equiv \int_0^\infty\left(\frac{d\mu(t)}{dt}\right)^2\left[1+(h-1)F_2(t)\right]dt.
\end{align}
With our piecewise constant history $(\boldsymbol y, \boldsymbol z)$ the derivatives become Dirac delta functions at the change points, and it is straightforward to show (using notation as in section \ref{sec:model:pcsws}) that these become
\begin{align}
\rho_\eta(\boldsymbol y, h) &= \|\Gamma(\boldsymbol y, h) \boldsymbol y\|_2^2,\\
\rho_\mu(\boldsymbol y, \boldsymbol z, h) &= \|\Gamma(\boldsymbol y, h) \boldsymbol z\|_2^2.
\end{align}
where

\emph{this is wrong, see notes!}

\begin{align}
\label{eqn:Gamma}
\Gamma(\boldsymbol y, h) \equiv&
\begin{bmatrix}
\sqrt{h+(1-h)\sum_2^n A_{2,j}u_0^{\binom{j}{2}}} &     &        &                       \\
  & \sqrt{h+(1-h)\sum_2^n A_{2,j}(u_0u_1)^{\binom{j}{2}}} &        &                       \\
  &     & \ddots &                       \\
  &     &        & \sqrt{h+(1-h)\sum_2^n A_{2,j}\left(\prod_{i=1}^{m-1}u_i\right)^{\binom{j}{2}}}
\end{bmatrix}\\
&\times
\begin{bmatrix}
1  &      &        &             &       \\
-1 & 1  &        &             &       \\
     & -1 & 1    &             &       \\
     &      & \ddots & \ddots      &       \\
     &      &        & -1 & 1
\end{bmatrix}.
\end{align}
Note that if $\boldsymbol y$ is fixed $\Gamma(\boldsymbol y)$ is a constant and thus a proper Tinkhonov matrix.

Our regularized loss function is then
\begin{align}
L(\boldsymbol y, \boldsymbol z) &= - \ell(\boldsymbol y, \boldsymbol z) + \lambda_\eta \rho_\eta(\boldsymbol y) + \lambda_\mu \rho_\mu(\boldsymbol y, \boldsymbol z)\\
&= -\boldsymbol x^\intercal\log(C M(\boldsymbol y) \boldsymbol z) + \boldsymbol 1^\intercal C M(\boldsymbol y) \boldsymbol z + \lambda_\eta \|\Gamma(\boldsymbol y, h) \boldsymbol y\|_2^2 + \lambda_\mu \|\Gamma(\boldsymbol y, h) \boldsymbol z\|_2^2
\end{align}
where $\lambda_\eta$ and $\lambda_\mu$ are parameters tuning the regularization strength, and we've dropped the constant term from $\ell$.

@@

>>>>

Following \cite{Rosen2018-bb} (appendix proof of Prop.\ (1)) let $R_\eta(t) \equiv \int_0^t\frac{dx}{\eta(x)}$, and substitute $\tau \equiv R_\eta(t)$ in \ref{eqn:d} to give
\begin{equation}
\label{eqn:d2}
d_j = \int_0^\infty \exp\left[-\binom{j}{2}\tau\right] \tilde\eta(\tau)\tilde\mu(\tau)dt,
\end{equation}
where $\tilde\eta(\tau) \equiv \eta(R^{-1}(\tau))$ and $\tilde\mu(\tau) \equiv \mu(R^{-1}(\tau))$.
We consider piecewise constant $\eta$ and $\mu$ on $m$ common epochs $[t_0, t_1), [t_1, t_2),\dots, [t_{m-1}, t_m)$ where $0=t_0 < t_1 < \dots < t_{m-1} < t_m=\infty$ (not to be confused with the coalescent time realizations in section \ref{sec:model:xi}).
We take the epochs as fixed parameters, and in practice make them dense so as to approximate infinite dimensional histories.
Let $(y_1,\dots,y_m)$ denote the constant population size $\eta(t)$ during each epoch, and let $(z_1,\dots,z_m)$ denote the constant mutation rate $\mu(t)$ during each epoch.
Let $u_l \equiv \exp(-(t_l-t_{l-1})/y_l)$ for $l=1,\dots,m$. %, and $u_0\equiv 1$.
With this we can follow the proof of Prop.\ (1) in \cite{Rosen2018-bb} mutatis mutandis to arrive at
\begin{equation}
\label{eqn:d3}
\boldsymbol d = M(\boldsymbol y) \boldsymbol z
\end{equation}
where
\begin{equation}
\label{eqn:M}
M(\boldsymbol y) \equiv
\begin{bmatrix}
1 &             &        &                       \\
  & \frac{1}{3} &        &                       \\
  &             & \ddots &                       \\
  &             &        & \frac{1}{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
1       & u_1                & \hdots & \prod_{i=1}^{m-1}u_i               \\
1       & u_1^3              & \hdots & \prod_{i=1}^{m-1}u_i^3             \\
\vdots  & \vdots             & \ddots & \vdots                                \\
1       & u_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{m-1}u_i^{\binom{n}{2}}
\end{bmatrix}
\begin{bmatrix}
1  &      &        &             &       \\
-1 & 1  &        &             &       \\
     & -1 & 1    &             &       \\
     &      & \ddots & \ddots      &       \\
     &      &        & -1 & 1
\end{bmatrix}
\begin{bmatrix}
y_1 &     &      &             &       \\
    & y_2 &      &             &       \\
     &     & y_3 &             &       \\
     &      &    & \ddots      &       \\
     &      &        &  & y_m
\end{bmatrix}.
\end{equation}
Note that the $(n-1)\times m$ matrix $M(\boldsymbol y)$ is a nonlinear function of the demographic history $\boldsymbol y$ because the $u_l$ are nonlinear functions of $\boldsymbol y$.

@@

\subsection{Poisson random field log likelihood}\label{sec:model:ell}

@@

The inverse problem \eqref{eqn:xivec2} is ill-posed in general, so many very different histories can be equally consistent with the data \citep{oscillation paper? Yun's other papers?}.
To penalize oscillatory solutions we introduce regularizers on the history derivatives.
Intuitively, we know the SFS can't contain any information about the history beyond the TMRCA $T_2$, since mutations that occurred before then will not be segregating in the sample.
Thus, we will find it useful to penalize complexity in the history more heavily at times that are more probably ancestral to the TMRCA.
To do this we should like to use the CDF of the TMRCA to define a ramp of weights.
At times for which the CDF is near unity, the TMRCA has probably occurred, so we should penalize heavily because the data are not informative about the history at this time.
At times for which the CDF is small, the TMRCA has probably not yet occurred, and we should penalize less heavily.
From \eqref{eqn:pi} and \eqref{eqn:r} the CDF is
\begin{align}
F_2(t) &\equiv 1 - \sum_{j=2}^n A_{2,j}r_2(t)\\
&= 1 - \sum_{j=2}^n A_{2,j}r_2(t)
\end{align}

We define $L_p$ derivative regularizers for arbitrary histories $\eta(t)$ and $\mu(t)$ to have unit weight when the CDF is zero, and approach weight $h$ as the CDF approaches unity:
\begin{align}
\rho_\eta[\eta(t); h, p] &\equiv \int_0^\infty\left|\left(1+(h-1)F_2(t)\right)\frac{d\eta(t)}{dt}\right|^p dt,\\
\rho_\mu[\eta(t), \mu(t); h, p] &\equiv \int_0^\infty\left|\left(1+(h-1)F_2(t)\right)\frac{d\mu(t)}{dt}\right|^p dt.
\end{align}
With our piecewise constant history $(\boldsymbol y, \boldsymbol z)$ the derivatives become Dirac delta functions at the change points, and it is straightforward to show (using notation as in section \ref{sec:model:pcsws}) that these become
\begin{align}
\rho_\eta(\boldsymbol y, h, p) &= \|\Gamma(\boldsymbol y, h) \boldsymbol y\|_p^p,\\
\rho_\mu(\boldsymbol y, \boldsymbol z, h, p) &= \|\Gamma(\boldsymbol y, h) \boldsymbol z\|_p^p.
\end{align}
where
\begin{align}
\label{eqn:Gamma}
\Gamma(\boldsymbol y, h) \equiv &\left(
\begin{bmatrix}
h-1 &   &   &        & \\
    & h &   &        & \\
    &   & h &        & \\
    &   &   & \ddots & \\
    &   &   &        & h
\end{bmatrix}
+ (1-h)\times
\text{diag}\left(
\begin{bmatrix}
A_{2,2} & A_{2,3} & \hdots & A_{2,n}
\end{bmatrix}
\times
\begin{bmatrix}
1       & u_1                & \hdots & \prod_{i=1}^{m-1}u_i               \\
1       & u_1^3              & \hdots & \prod_{i=1}^{m-1}u_i^3             \\
\vdots  & \vdots             & \ddots & \vdots                                \\
1       & u_1^{\binom{n}{2}} & \hdots & \prod_{i=1}^{m-1}u_i^{\binom{n}{2}}
\end{bmatrix}\right)\right)\\
&\times
\begin{bmatrix}
1  &      &        &             &       \\
-1 & 1  &        &             &       \\
     & -1 & 1    &             &       \\
     &      & \ddots & \ddots      &       \\
     &      &        & -1 & 1
\end{bmatrix}.
\end{align}
% Note that if $\boldsymbol y$ is fixed $\Gamma(\boldsymbol y)$ is a constant and thus a proper Tinkhonov matrix.

Our regularized loss function is then
\begin{align}
L(\boldsymbol y, \boldsymbol z) &= - \ell(\boldsymbol y, \boldsymbol z) + \lambda_\eta \rho_\eta(\boldsymbol y, h, p) + \lambda_\mu \rho_\mu(\boldsymbol y, \boldsymbol z, h, p)\\
&= -\boldsymbol x^\intercal\log(C M(\boldsymbol y) \boldsymbol z) + \boldsymbol 1^\intercal C M(\boldsymbol y) \boldsymbol z + \lambda_\eta \|\Gamma(\boldsymbol y, h) \boldsymbol y\|_p^p + \lambda_\mu \|\Gamma(\boldsymbol y, h) \boldsymbol z\|_p^p
\end{align}
where $\lambda_\eta$ and $\lambda_\mu$ are parameters tuning the regularization strength, and we've dropped the constant term from $\ell$.

\subsection{$p=1$}

To optimize with $p=1$ we use the alternating direction method of multipliers (ADMM) of \cite{Boyd2011-dm}, which allows us to handle the non-differentiable penalty term.

@@

\section{Acknowledgements}\label{sec:ack}
Jean Feng suggested ADMM (if we end up using that).
